---
title: "Hurricane Code"
output: 
  pdf_document: default
  html_document: default
date: "2024-05-09"
---

```{r}
# libraries installed
library(tidyverse)

theme_set(theme_bw()) # optional

# dataset storms
storms
```

# Sections

### Section 1: Frequency of status types among the years

notes:

1)  Tropical storms are the most frequent type while tropical wave is least frequent

```{r}

storms %>% 
  rename(Classifications = status)  %>% 
  ggplot(aes(year)) + geom_bar(aes(fill = Classifications), position = "fill", alpha = 0.9) + theme_bw() +
  theme(legend.background = element_rect(fill = "white", linetype = "solid", color = "black")) +
  labs(title = "Different Classifications types by Year",
       subtitle = "Percentage of Different Storms by Year",
       x = "Years",
       y = "")

# Total count of each status 

storms %>% 
  count(status, sort = TRUE)
  

```

### Section 2: Difference of latitude among different status

notes:

1)  Tropical waves appear to have the least variation among others.
2)  Extratropical appears to have the highest variation among the others.

```{r}

# Graph
storms %>% 
ggplot(aes(x = lat)) + geom_density(aes(color = status, fill = status), alpha = 0.2) +
  labs(title = "Difference of Latitude",
       subtitle = "Among Different Storm Classifications",
       x = "Latitude",
       y = "Density")

# By status
storms %>% 
  ggplot(aes(x = lat, color = status)) + geom_density(aes(fill = status), alpha = 0.2) + facet_wrap(~status, nrow = 3) +
  labs(title = "Additional Latitude Visualization",
       subtitle = "Among Different Storm Status",
       x = "Latitude",
       y = "Density") 

# Shows mean and variance on lat for each status
storms %>% 
  group_by(status) %>% 
  summarise(mean_lat = mean(lat),
            sd_lat = sd(lat),
            var_lat = var(lat))

```

### Section 3: Difference of longitude among different storm classfications

```{r}

# graph
ggplot(storms, aes(x = long, color = status)) + geom_density(aes(fill = status), alpha = 0.2) + theme_bw()

# By status
ggplot(storms, aes(x = long, color = status)) + geom_density(aes(fill = status), alpha = 0.3) + facet_wrap(~status, nrow = 3) 

# Table that shows mean and variance for long on each status
storms %>% 
  group_by(status) %>% 
  summarise(mean_long= mean(long),
            sd_long = sd(long),
            var_long = var(long))

```

### Section 4: Wind

Notes:

1)  Wind is right skewed.
2)  Hurricanes have the highest mean for wind speeds while other low has the lowest.

```{r}

# Wind across all years
storms %>% 
  ggplot(aes(wind)) + geom_density(aes(fill = "red"), alpha = 0.7) + theme_bw()

# Wind by status
storms %>% 
  ggplot(aes(wind)) + geom_density(aes(color = status, fill = status), alpha = 0.3) + facet_wrap(~status) 

# Wind by status in one graph
storms %>% 
  ggplot(aes(wind)) + geom_density(aes(color = status, fill = status), alpha = 0.2)

# Box plots for different status types
storms %>% 
  mutate(status = fct_reorder(status,wind)) %>% 
  ggplot(aes(wind, status)) + geom_boxplot(aes(fill = status), outlier.size = 0.5) + 
  theme_bw() +
  labs(
    title = "Box and Whisker Plots for Wind",
        subtitle = "Among Storm Classifications",
        x = "Wind",
        y = "Classifications"
  )

# Table that displays the mean and variance on wind for each status
storms %>% 
  group_by(status) %>% 
  summarise(mean_wind = mean(wind),
            sd_wind = sd(wind),
            var_wind = var(wind))





```

### Section 5: Pressure

Notes:

1)  Pressure is left skewed
2)  The mean for pressure is very close among: Disturbance, other low, subtropical depression, tropical depression and tropical wave

```{r}

# Pressure across all years
storms %>% 
  ggplot() + geom_density(aes(pressure), fill = "blue", alpha = 0.5) + theme_bw()

# By status
storms %>% 
  ggplot() + geom_density(aes(pressure, color = status, fill = status), alpha = 0.5) + facet_wrap(~status)

# All in one graph
storms %>% 
  ggplot() + geom_density(aes(pressure, color = status, fill = status), alpha = 0.3)

# Box plots for different status types
storms %>% 
  rename(Classification = status) %>% 
  filter(pressure > 920) %>% 
ggplot(aes(pressure, Classification)) + geom_boxplot(aes(fill = Classification), outlier.size = -3) + labs(
        title = "Box and Whisker Plots for Pressure",
        subtitle = "Among Storm Classifications",
        x = "Pressure",
        y = "Classifications"
)


# Table that displays the mean and variance on pressure for each status
storms %>% 
  group_by(status) %>% 
  summarise(mean_pressure = mean(pressure),
            sd_pressure = sd(pressure),
            var_pressure = var(pressure))




```

### Graph 10: Pressure vs. Wind by Status

```{r}

# Visual
storms %>% 
  group_by(status) %>% 
  rename(Category = category) %>% 
  ggplot(aes(pressure,wind)) + 
  geom_point(aes(color = Category)) + 
  geom_jitter(aes(color = Category)) +
  theme(legend.background = element_rect(fill = "lightblue", linetype = "solid", color = "black")) + 
  theme(legend.position = "top") +
  geom_smooth(color = "red") +
  facet_wrap(~status, nrow = 3) +
  labs(title = "Pressure vs. Wind",
       subtitle = "Through 1975-2020",
       x = "Pressure in Millibars",
       y = "Wind in Knots")

  

```
